|

### 2021-08-22 dissertation/msc_dissertation.Rmd
Editing dissertation

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Defining the results space in @silberzahn2018 through model specification*
  |
  | Master's thesis - DRAFT
author: "Sebastian Ploner, sploner1@sheffield.ac.uk, University of Sheffield"
linestretch: 1.5
output:
    pdf_document:
        includes:
            in_header: header.tex
        fig_caption: yes
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
  - \usepackage[ruled,vlined]{algorithm2e}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font={stretch=1}}
---

@@

knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H')

@@

# Introduction
## Theoretical framework
     The Covid-19 pandemic has reaffirmed the need for sound scientific research to inform decision-making on a scientific and societal level [@collins2021]. Rigorous research builds upon a systematic and well-reasoned approach to solving a research problem. Based on available literature, a falsifiable research question is defined and a corresponding hypothesis is developed. An appropriate study is then designed and conducted. To draw inferences from the gathered data, statistical models are developed and the effect of each variable is assessed. Every one of these steps is influenced by the decisions researchers make, which are known as *researcher degrees of freedom* [RDF, @simmons2011; @wicherts2016]. Usually there are many feasible analytical strategies to answering a research question and none of them are inherently right or wrong  [@carp2012]. This often creates uncertainty, for example, about what covariates to include and how to model them, which in turn leads to inconsistent findings [@patel2015; @ioannidis2008]. Recently, efforts have been made to better understand the scope of variation induced by different analytical strategies.

 These efforts include crowdsourcing approaches to data analysis [@silberzahn2018; @botvinik-nezer2020]. Its premise is to have a large number of researchers team up into to smaller, independent groups to investigate the same research question based on the same dataset. For instance, @silberzahn2018 had 29 teams investigate the effects of a football player's skin colour on the odds of being sent off the field. The variation between the analytical strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect. The authors also controlled for researchers' prior believes and experience as well as peer-rated analysis quality, none of which accounted for the variation of results. @botvinik-nezer2020 made similar observations. Crowdsourcing projects excels at emphasising the substantial impact of different analytical strategies, but they are also extremely time and personnel intensive. The two mentioned studies lasted between 2 to 3+ years, and included 61 and 180 analysts, respectively. Additionally, despite covering a more diverse set of analytical strategies than a single conventional analysis, crowdsourcing is limited by the number of participating teams. It, therefore, still only covers a few selected strategies, not all possible ones.

 @silberzahn2018 and @botvinik-nezer2020, therefore, suggest using multiverse analysis (also known as specification-curve analysis) as an alternative to crowdsourcing. This approach requires identifying and running all plausible analytical strategies. Plausible strategies are defined as statistically valid, non-redundant, tests appropriate to the research question. Their aggregated results are then used to make inferences about the research question. Despite being statistically more complex and computationally more intense, this approach has the advantages of only needing one, or ideally a couple, researchers. Moreover, a researcher's inherent strategy bias is neutralised and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited for assessing the scope of variation induced by different analytical strategies. However, one aspect has been, to my knowledge, neglected thus far. To properly assess the scope of variation it needs to be contextualised. That is, the space in which it can vary needs to be defined in order to understand to what degree the scope covers the space. The concept of statistical association provides a suitable analogy. Covariation and correlation both indicate the direction of relationship (positive, null, negative), however, only correlation allows for assessing the association's strength. Correlation is the standardised version of covariation, and only knowing the its limits gives the correlation coefficient meaning. The same principle applies to analyses on a broader level.

## Current project
     Taken together, researchers make many analytical decisions throughout conducting a study introducing so-called *researcher degrees of freedom*. Crowdsourcing analysis presents a strategy to understand their induced variation, but is impractical due to its time and personnel requirements, and is still limited by the number of participating teams. Multiverse analysis present a promising, low-personnel, but computationally intense alternative to assessing the scope this variation. Despite observing substantial variation in crowdsourced data analysis projects, little is known about the extend to which these approaches cover the full range of possible results. Determining the full results space and its underlying mechanisms is an important exercise to understand the impact of analytical strategies. To my knowledge no previous study has attempted to define and understand a study's entire results space. In this project, running all possible analytical strategies, I, therefore, aim to define and understand the results space of @silberzahn2018. By doing so, I hope further the understanding of how an analytic strategy affects its result.

# Analysis
## Analysis plan
     The project's objective is to define the results space of the @silberzahn2018 to draw conclusions about the mechanisms of multiverse analysis. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analytical strategy. Hence, it is closely related to assessing an effect's robustness. Robustness refers to an effect's consistency under different model specifications. @patel2015 developed a standardised approach to assessing an effect's robustness which they termed assessing the *vibration of effect*. I will therefore use this standardised approach to define the results space. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018 the analysts used numerous different statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams there 15 covariates used. This gives $2^n$ i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation gives a total of $29 * 2^{15} = 950,272$ combinations to run. This number exceeds the project's scope, hence, it needs to be reduce to a more manageable count. I am, therefore, focusing on one modelling approach. In particular, on the approach that produced the median outcome of all analyses. The median, being the middle number of any set of values, is a reasonable starting point in order to estimate the results space. Nevertheless, even focusing on one analytical approach still leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations I will first run a sample of 200, followed by a sample of 1,000 combinations. If there's a substantial difference between the median and the spread I will run more models, if not, I will make the assumption that 1,000 reasonably estimates all ~33k combinations.

 In @silberzahn2018 the median outcome was produced by @team23 which will hereinafter be referred to as *team 23* due to it's designated team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of this project will be replicating the transformation and the analysis. Replicating other researchers analyses has the benefit of checking their work and, if results are indeed replicated, increasing confidence in them. It also ensures that this project has the same starting point as *team 23* did. Given the team made all their scripts publicly available I expect this step to be straightforward. Next is drawing a random sample of covariates, without replacement. "Without replacement" ensures all covariate combinations are unique in the sample. The covariates will be appended to the base (or core) variables. Base variables are those variables that are primarily assessed to answer the research question. In this case whether a football player's skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: "skin tone *X* implicit bias" and "skin tone *X* explicit bias". (The variables are described in the "data" section.) Hence, all models will have the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
While running each model, the relevant statistics will be extracted. Those are the coefficient (or effect) of skin tone, its standard error, test-statistic and p-value. Just as *team 23* did I will then calculate the 95% confidence intervals (CI). The estimates and their CIs are then transformed to odds ratios (OR) through exponentiating them to the power of two. Odds ratios quantify the strength of association between two variables. If greater than one the dependent variable is more likely to occur given the independent variable, if lower than one it less likely.

 These odds ratios will be visualised to assessed through a conventional specification curve plot, a rain-cloud plot and scatter plot. The following describes the three graphs in more detail:

(i) The specification curve plot was developed to be a descriptive plot i.e., raw data is being plotted without any aggregation done. Its objective is to describe the results space while allowing the reader to identify the specified covariates for each outcome [@simonsohn2020]. It therefore has two vertically stacked components. The top part shows a sorted scatter plot. Its horizontal axis shows the models specifications and its vertical axis displays the outcome measure. The scatter points are sorted from lowest to highest outcome measure. This way the lowest outcome measure is on the bottom left corner and the highest in the top right one. The bottom part of the plot represents a table. As in the top plot the horizontal axis holds the model specifications, the vertical axis lists the covariates. This arrangement allows each cell to specify the presence or absence for each covariate in a given specified model. For the top and bottom plot to work together it is imperative that the horizontal axes are identically arranged. If so, for every point in the top plot (i.e., for every outcome) the specified covariates can be seen in the bottom plot. For an example check Figure \ref{fig:sca_plot}. As the current project has 1,000+ specifications it becomes very hard to identify specific covariates given the space limitations. Hence, I to had come up with different more intuitive ways of visualisation.

(ii) The rain-cloud plot was developed by @allen2021. Its objective is to give an unbiased, transparent view on the raw data. It therefore combines a density, box and scatter plot. If stacked vertically from top to bottom, respectively, they look like a cloud with rain drops, hence, its name. The advantage of this plot is to be able to assess the raw data including key statistics (median, 25th and 75th quartiles and CIs) as well as the probability density all in one succinct plot. The current project seeks to define and describe a results space the rain-cloud plot is, therefore, the ideal choice of graphs. (For an example check the results section.) Unfortunately, the plot does not allow to identify the covariates giving rise to the outcomes or their specified effects. As the project also seeks to understand the mechanisms of multiverse analysis it important to get insights about the specified covariates effects. Therefore a third plot is developed which is to be assessed in tandem with this plot.

(iii) The third and final plot is a sorted scatter plot. While there is nothing special about the plot itself its content is important. Given the very high number of models that are run, it does not make sense to assess each model and its effects in isolation. Instead the effects for each covariates need to be aggregated. Hence, the goal of this plot is to visualise the specified effect for each covariate. Prior to doing so, the covariates first need to be aggregated and their specified effects need to be calculated. To this end, each covariate is recoded into a binary factor: included in the model yes/no. This newly defined factors are than used as an independent variables in an ANOVA. The previously calculated odds rations are used as dependent variables. The estimates of the fitted model are the specified effect for each covariates. These are visualised similarly to the top part of the specification curve plot. The covariates are on the horizontal axis, will the odds ratios are on the vertical axis. Here too, the lowest outcome is in the bottom left corner and the highest outcome in the top right one. Additionally the point are coloured indicating a statistically significant effect (or non-significant).

 Taken together, the current project's objective is to define the results space of @silberzahn2018 in order to make inferences about the underlying mechanisms of multiverse analysis. The results space is created by running all possible model and covariate combinations. Applied to @silberzahn2018 this means running roughly 1M models. Due to time and computational limitations, I am focusing on the model that produced the median outcome of all analytical strategies. Hence, I will run a mixed-model logistic regression with 1,000 randomly sampled covariate combinations. For each model relevant statistics will extracted and visualised. The conventional specification curve plot does not deal well with a large number of model specifications. Therefore, I will produce a rain-cloud and scatter plot, which visualise the results space (and its distribution) as well as the specified covariate effects, respectively. For the latter, I will first have to run an additional statistical model to calculate the specified effects. These graphs will give insight into the results space, its building blocks and ultimately into the mechanisms of multiverse analysis.

## Code
     This section provides a brief overview of the code and explains the reasoning behind it. As mentioned in the analysis section the first step was replicating the data transformation and analysis of *team 23* [@team23]. The team made their project folder publicly available. It contains three scripts relevant to this project: the data exploration, transformation and analysis. After duplicating their project folder on my local hard drive all scripts run without issues (only the working directories needed adjustment). The data exploration included the reasoning behind transforming the data and some cleaning. The data transformation restructured the data into a more intuitive format (more details on the topics of exploration and transformation are in the data section). The analysis script prepared the data by assigning variables classes (factors, numerical or boolean) and standardised a few variables i.e., they were centred around the mean. Finally, the models were specified. The team ran both frequentist and Bayesian models, however, this project covers the frequentist approach, only.

 My code is based on @haessler2020 and @patel2015. The rough structure of the code is shown in the pseudocode table (Figure X). Here, I am briefly motivating two analytical choices I made. For a closer look please check the commented code itself:

(i) Choosing the function to run the statistical model. *Team 23* ran a mixed-model logistic regression. This model has two relevant components to this explanation: fixed effects and random effects. Fixed effects are those that are consistently observed in different situations because the construct is of direct interest to the research question. In this case, it is, for example, skin tone. Independent of the player, the game or the league skin tone is always observed. Their counter parts are random effects which change between situations. A player, for example, is just one "unit" a measurement, and is himself not directly relevant to answering the research question. The function used by *team 23* requires to specify a random effect. However, given the project seeks to sample from all possible covariate combinations, a random effect is not always included. Hence, the "non-random" counter part to this function needs to be utilised. For this project, it was important to check that both functions use the same estimation method (maximum likelihood or restricted maximum likelihood estimation) to calculate the statistics. It was important to check to ensure the outcomes are comparable.

(ii) No multiple comparison. This is based on the fact that in statistics null hypotheses are one rejected if the probability of finding a false positive is below 5% (known as the significance level, $\alpha$). The more statistical tests a run, the more likely it is that one of those tests is indeed a false positive. Hence, the are techniques that account of this issue. It is usually good practice to account for those "multiple comparisons". Here, I refrained from doing so though. This project seeks to simulate multiple researchers running different models. Therefore, for the the purpose of this project all models are run independently by individual researchers. Those would not have knowledge of the other analyses and, hence, would not account for them. To main the highest possible ecological validity, I therefore did not account of multiple comparisons.

\begin{algorithm}
\caption{Pseudocode overview--in progress}
\DontPrintSemicolon
\SetAlgoLined
\KwData{Prepared data based on team 23 script}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output} 
$variables$ $\leftarrow$ Define dependent, base variables and covariates \\
$specifications$ $\leftarrow$ Use $variables$ to create matrix containing all possible covariate combinations \\
$formula$ $\leftarrow$ Paste $specifications$ by row and append as column to $specifications$ \\
$formula.ranef; formula.ef$ $\leftarrow$ Separate formulas based on including a random-structure \\
$samlple.ranef; sample.ef$ $\leftarrow$ Random, without replacement, sample from $formula.ranef; formula.ef$
\BlankLine
 initialization\;
  \While{not at end of this document}{
    read current\;
    \eIf{understand}{
      go to next section\;
      current section becomes this one\;
      }{
      go back to the beginning of current section\;
      }
    }
\KwResult{List containing model statistics}
\end{algorithm}

 The code is written in R (Version 1.4.1103) on macOS Big Sur (Version 11.4) and can be retrieved from my [GitHub repository](https://github.com/sebastianplnr/msc_dissertation_project). The code from *team 23* [@team23] can be retrieved from their [OSF repository](https://osf.io/akqt4/). Following R packages were used *here 1.0.1* [@here], *data.table 1.14.0* [@datatable], *tidyverse 1.3.1* [@tidyverse], *lme4 1.1-27.1* [@lme4], *pbmcapply 1.5.0* [@pbmcapply], *PupillometryR 0.0.3* [@PupillometryR] and *cowplot 1.1.1* [@cowplot].

## Data description and preparation
     The data was retrieved from @silberzahn2018. It contains information about football players, their encounters with referees and the received cards (yellow, yellow/red and red). Moreover, a player's position, age, club, league country, victories, ties, defeats and goals. In addition, a skin tone rating based on two independent raters was included. The referees are numerically coded to protect their identify. The referees origin countries are also included as well as an implicit and explicit racism bias scores for their respective countries. The exploratory data analysis (EDA) of *team 23* showed that each row of the dataset represents a unique player-referee combination listing all their encounters as well as the couple's total number of received/assigned cards. The team states that it preferred a different data format. More specifically, a format where each player-referee encounter is reflected by one row. This way each encounter had a maximum of one red card. To achieve this format the data had to be "disaggregated" i.e., transformed. Further EDA showed that receiving a red card is highly unlikely, 0.008%, i.e., data is highly skewed. This property informed the teams decision to use a type of logistic regression (a statistical modelling technique equipped to deal with skewed/binary distributions). Finally, the team excluded all referee that did have at least 22 encounters. Their reasoning was a game includes (at least) 22 players, 11 players per team, if this were not the case referee most likely did not officiate in one for the four leagues of interest (England, Germany, Italy and Spain). This step excluded roughly 66% of the referees but retained 97.4% of the cases. The final dataset used for the current project contained 335,537 observations and 19 variables. Figure \ref{fig:var_plots} shows the properties of the variables relevant to current project's analysis \newline

```{r var_plots, echo = FALSE, fig.width = 8, fig.align = "center", fig.cap = "A) Showing the strong skew of receiving no red to receiving a red card. B) Showing most players had a lower skin tone rating i.e., most players were on the whiter side. C) Showing there is not much variation in implict racial scores. D) Showing more variation in explitic reacial bias scores. The latter two were centred around the mean, hence, zero does not mean there is no bias."}

@@

# Results
## Replicating *Team 23* analysis
     As mentioned in the above section the first steps were replicating the *team 23* analysis using their provided scripts. The results are reproducible without any adjustment. Skin tone has a significant effect on the odds of being send off the field ($OR = 1.311$, 95%CI [1.099, 1.563], $p = 0.003$). This means while keeping all other variables constant for every unit increase in skin tone rating (darker skin tone), the odds of being send off the field increase by roughly 131%. The interaction terms of skin tone and implicit racial bias ($OR = 0.004$, 95%CI [0.000, 23.259], $p = 0.211$) as well as skin tone and explicit bias ($OR = 1.837$, 95%CI [0.493, 6.848], $p = 0.365$) are both non-significant. In accordance with the original analysis, there are also significant differences between leagues and positions as well as implicit racial bias scores. Explicit bias scores are on verge of being non-significant. This analysis only aimed at testing reproducibility, which proved to be true, further results details are forfeited.

>>>>

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Assessing results space and structure of @silberzahn2018 through model specification*
  |
  | MSc Psychological Research Methods with Data Science
  | Dissertation
author: "Sebastian Ploner, sploner1@sheffield.ac.uk, University of Sheffield"
output:
  pdf_document:
    includes:
      in_header: header.tex
    fig_caption: yes
  word_document: default
linestretch: 1.5
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
- \usepackage[ruled]{algorithm2e}
- \usepackage{float}
- \usepackage{caption}
- \captionsetup[figure]{font={stretch=1}}
---

@@

knitr::opts_chunk$set(echo = FALSE, fig.pos = 'H')

@@

# Introduction
## Theoretical framework
     The Covid-19 pandemic has reaffirmed the crucial relevance of sound scientific research for political and societal decision-making [@collins2021]. Rigorous research builds upon a systematic and well-reasoned approach to solving a research problem. Based on current knowledge, researchers define a research question and develop a hypothesis. To test the hypothesis, they design and conduct a research study that yields data. To draw conclusions from the data, researchers apply statistical models and assess how different variables have influenced the data. Each of these steps is influenced by the researchers’ decisions, which are known as *researcher degrees of freedom* [RDF, @simmons2011; @wicherts2016]. In most cases, there is not only one but many feasible analysis strategies to answer a research question [@carp2012]. This ambiguity often creates uncertainty and inconsistency. Specifically, researchers are often uncertain about which covariates to include and how to model them, which leads to inconsistent findings [@patel2015; @ioannidis2008]. Recently, efforts have been made to better understand how different analysis strategies influence research results.

 Crowdsourcing is one approach to better understand the influence of analysis strategies on research findings [@botvinik-nezer2020; @silberzahn2018]. A large number of researchers team up into smaller, independent groups to investigate the same research question based on the same dataset. For instance, in @silberzahn2018, 29 teams investigated the effects of a football player’s skin colour on the odds of being sent off the field. The variation between the analysis strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect of skin colour on the risk of being sent off the field. The authors also controlled for researchers’ prior believes and experience as well as peer-rated analysis quality, but these factors did not account for the variation of results. @botvinik-nezer2020 made similar observations. Such crowdsourcing approaches are well suited to show how different analysis strategies influence research findings, but they are extremely time and resource consuming. For instance, it took two and more than three years and 61 and 180 analysts, respectively, to perform the two studies mentioned above. Furthermore, although crowdsourcing approaches can cover more analysis strategies than a single conventional analysis, they are limited by the number of teams. They, therefore, cover a specific selection but not all possible strategies.

 Another approach to assess the effects of analysis strategies on research findings is multiverse analysis (also known as specification-curve analysis, [@simonsohn2020]. This approach has been proposed by @botvinik-nezer2020 and @silberzahn2018 as an alternative to crowdsourcing. In this approach, all reasonable analysis strategies are identified and performed. *Reasonable strategies* are defined as all statistically valid, non-redundant tests that are appropriate for the research question. The aggregated results of all reasonable analysis strategies are used to make inferences about the research question. Although the approach is statistically complex and computationally intense, it can be performed by one or a few researchers. Moreover, a researcher’s inherent strategy bias is neutralised, and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited to better understand how different analysis strategies influence research results.

## Current project
     Researchers make analytical decisions that introduce *researchers degrees of freedom*. Crowdsourcing analysis and multiverse analysis are promising tools to assess the variation of results induced by analysis strategies. However, these approaches do not assess the full range of results obtained by all possible analysis strategies but usually focus on a subset of reasonable analysis strategies.  Therefore, it is unclear how the spaces of all reasonable and all possible analysis strategies relate to each other. In particular, it is unknown whether the space of all reasonable results covers the full range of possible results or whether it covers only a narrow sub-space of possible results. In this project, all reasonable and all possible analysis strategies were performed to define and understand the results space of @silberzahn2018. Furthermore, ways to visualise the space and its structure intuitively and succinctly were proposed. The project is, thus, intended to further the basic understanding of how analysis strategies influence research results and to provide tools for further studies on this topic.

# Analysis
## Analysis plan
     The project's objective was to define the space and the structure of all possible results of @silberzahn2018. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analysis strategy. Hence, it is closely related to assessing the robustness of an effect. Robustness refers to the consistency of an effect under different model specifications. @patel2015 developed a standardised approach to assess an effect's robustness. I used this standardised approach to define the results space. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018, the analysts used numerous statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total, there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams, 15 covariates were used, resulting in $2^n$, i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation yields a total of $29 * 2^{15} = 950,272$ possible analysis strategies. As this number of strategies exceeds the current project's resources, it was reduced to a more manageable number. Therefore, the project focused on the modelling approach that produced the median outcome of all analyses. The median, being the middle number of a given set of values, is a reasonable starting point to estimate the results space. However, even focusing on one analysis approach leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations, a random sample of 1,000 analysis strategies were conducted.

 In @silberzahn2018, the median outcome was produced by @team23, which will hereinafter be referred to as *team 23* due to its team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of the present project was to replicate their transformation and analysis. Such replication can increase the confidence in the previous and the current approach. Moreover, it ensures that this project has the same starting point as *team 23*. As *team 23* made all scripts publicly available, this step was straightforward. The next step of the project was to define the results space and its structure. To this end, a random sample of covariates without replacement was drawn. “Without replacement” ensured that all covariate combinations were unique in the sample. The covariates were appended to the base (or core) variables. Base variables were those variables that were primarily assessed to answer the research question. In this case, the research question was whether a football player’s skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: “skin tone X implicit bias” and “skin tone X explicit bias.” (The variables are described in the “data” section.) Hence, all models had the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
The relevant outcome parameters of each model were extracted. These parameters were the coefficient (i.e., effect) of skin tone, its standard error, test statistic and p-value. Similar to *Team 23* I calculated the 95% confidence intervals (CI). The estimates and their CIs were transformed to odds ratios (OR) through exponentiating them to the power of two. OR quantify the strength of association between two variables. An OR greater than one indicates that the dependent variable is more likely to occur given the independent variable, if it is lower than one it is less likely to occur. Eight OR outliers were excluded from visualisation. Four approached infinity, one was in the lower million range, one was 150, and a final one was 4.6.

 To provide insights into the results space and its structure, the OR were visualised and assessed by three complementary plots, (i) a specification curve plot, (ii) a raincloud plot, (iii) and a scatter plot. The following describes the three plots in more detail:

(i) The specification curve is a descriptive plot that shows raw outcome data without any aggregation. Its objective is to describe the results space while allowing the reader to identify the model specifications for each outcome [@simonsohn2020]. It has two vertically stacked components. The upper component is a sorted scatter plot. Its horizontal axis lists the specifications, and its vertical axis displays the outcome measure. The data points are sorted from lowest to highest outcome measure. This way the lowest outcome measure is on the bottom left corner and the highest in the top right one. The lower component of the plot is a table. Similar to the upper component, the horizontal axis lists the specifications, while the vertical axis lists the covariates. This arrangement allows each cell to specify the presence or absence of each covariate in a given model specification. For the integration of information from the upper and lower components, the horizontal axes are identically arranged. Thus, for each point in the upper component (i.e., for every outcome), the specified covariates can be seen in the lower component (see Figure \ref{fig:sca_plot} for an example). However, with the more than 1,000+ specifications of the current project, it was hard to identify specific covariates. Hence, the specification curve was complemented with other plots.

(ii) The raincloud plot @allen2021 combines a probability density plot, a box plot, and a scatter plot to give an unbiased, transparent view of the raw data. If the three plots are stacked vertically from top to bottom, they look like an eponymous cloud with rain drops. The strength of the raincloud plot is that it visualizes the raw data, summary statistics (median, 25th and 75th quartiles and CIs) and the probability density in a single plot. As the current project seeks to define and describe a results space, the raincloud plot is an ideal tool (for an example check Figure \ref{fig:rain_cloud_plot}). However, the plot does not allow to identify which covariates cause which outcomes. As the project also sought to understand the structure of the results space, it was essential to gain insights into how specific covariates influenced the outcomes. Therefore a third plot which complemented and extended the raincloud plot was developed.

(iii) The third plot was a sorted scatter plot. The goal of this plot was to visualise the effects of each covariate. Given the large number of models, assessing all models and their effects individually did not make sense. Hence, the effects of each covariate were aggregated. To this end, each covariate was recoded into a binary factor: included in the model yes/no. These newly defined factors were then used as independent variables in an ANOVA. The previously calculated OR were used as dependent variables. The estimates of the fitted model were the specified effect for each covariate. These were visualised similar to the top part of the specification curve plot. The covariates were on the horizontal axis, while the OR were on the vertical axis. The lowest outcome was in the bottom left corner, and the highest was in the top right corner. Additionally, the colours of the points indicated the statistical significance of the effects.

 In summary, the current project aimed to define the space and structure of the results of @silberzahn2018. The space of possible results is defined by running all models with all possible covariate combinations. Applied to @silberzahn2018 this would result in about 1M models. Due to time and computational limitations, the project focused on the modelling approach that produced the median outcome of all analyses. Hence, a mixed-model logistic regression with 1,000 randomly sampled covariate combinations was run. For each model, relevant parameters were extracted and visualised. The specification curve is not ideal for large numbers of model specifications. It was, therefore, complemented and extended by raincloud and scatter plots, which visualise the results space and the specified covariate effects, respectively.

## Code
     This section provides a brief overview of the code and explains the reasoning behind it. First, the data transformation and analysis of *Team 23* [@team23] were replicated. The team made their project folder publicly available. It contains three scripts relevant to the current project: data exploration, transformation, and analysis. After duplicating their project folder on the local hard drive, all scripts ran without issues (only the working directories needed adjustment). The data exploration included the reasoning behind transforming the data and some cleaning. The data transformation restructured the data into a more intuitive format (more details on the topics of exploration and transformation are in the data section). The analysis script prepared the data by assigning variable classes (factors, numerical or boolean) and standardised a few variables, i.e. they were centred around the mean. Finally, the models were specified. The team ran both frequentist and Bayesian models. However, the current project focused on the frequentist approach.

 The code was based on @haessler2020 and @patel2015. The basic structure of the code is shown in the pseudocode table (see Algorithm \ref{alg:pseudocode}). Here, two important analytical decisions are briefly motivated. For more details, please refer to the commented code:

(i) Choosing the function to run the statistical model. *Team 23* ran a mixed-model logistic regression. This model has two relevant components: fixed effects and random effects. Fixed effects are consistently observed in different situations because the construct is of direct interest to the research question. In this case, it is, for example, skin tone. Independent of the player, the game or the league, skin tone is always observed. Their counterparts are random effects which change between situations. A player, for example, is just one “unit” of measurement and is herself/himself not directly relevant to answering the research question. The function used by *Team 23* requires specifying a random effect. However, given that the current project sought to sample from all possible covariate combinations, a random effect was not always included. Hence, the “non-random” counterpart to this function was used. Therefore, this project ensured that both functions used the same estimation method (maximum likelihood or restricted maximum likelihood estimation) to ensure that the outcomes are comparable.

(ii) No correction for multiple comparisons. In statistics, it is common practice to reject a null hypothesis if the probability of finding a false positive is below 5% (known as the significance level, $\alpha$). The more statistical tests are run, the more likely it is that a result of at least one of the tests is a false positive. Hence, it is good practice to account for those “multiple comparisons.” The current project, nevertheless did not correct for multiple comparisons. This project sought to simulate multiple researchers running different models. Those researchers would not know the other analyses and, hence, would not account for them. To maintain the highest possible ecological validity no correction for multiple comparisons was made. \newline

\begin{algorithm}
\caption{Outlining high-level structure of the code. "Ran" in e.g. "FormulaRanEf" refers to "random-structure". "Ef" refers to "effect".}
\label{alg:pseudocode}
\SetKwInOut{Output}{Output}
\textbf{data} $\leftarrow$ prepared data based on team 23 \\
  \BlankLine
  $Variables$ $\leftarrow$ Define dependent ($dv$), base variables ($basevar$) and covariates ($covar$) \\
  $Specifications$ $\leftarrow$ Use $Variables$ to create matrix containing all possible covariate combinations \\
  \BlankLine
  $Formula$ $\leftarrow$ Paste $Specifications$ by row and append as column to $Specifications$ \\
  $FormulaRanEf; FormulaEf$ $\leftarrow$ Separate formulas based on including a random-structure \\
  \BlankLine
  $SampleRanEf; SampleEf$ $\leftarrow$ Unique, random sample from $FormulaRanEf; FormulaEf$ \\
  \BlankLine
  $ExtractorFunction$ $\leftarrow$ Extracting model statistics and store them in object \\
  \BlankLine
  $OutputsRanEf$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glmer($dv$, x = $basevar + SampleRanEf$)))} \\
  $OutputsRanEfEF$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glm($dv$, x = $basevar + SampleEf$)))} \\
  \BlankLine
  $ModelOutputs$ $\leftarrow$ Merge $OutputsRanEf$ and $OutputsEF$ \\
  $ModelOutputs$ $\leftarrow$ Compute odds ratios and confidence intervals \\
  \BlankLine
  $JoinedData$ $\leftarrow$ \textbf{LeftJoin($Specifications$, $ModelOutputs$, by = $Formula$)} \\
  \BlankLine
  \Output{Rain cloud plot}
  \BlankLine
  $CovariateEffects$ $\leftarrow$ \textbf{LinearModel($OddsRatios$, x = $Specifications$, data = $JoinedData$)} \\
  \BlankLine
  \Output{Covariate effects plot}
  \Output{SCA plot, sorted by $CovariateEffects$}
  \BlankLine
  \Output{Volcano plot}
\end{algorithm}

 The code is written in R (Version 1.4.1103) on macOS Big Sur (Version 11.4) and can be retrieved from my [GitHub repository](https://github.com/sebastianplnr/msc_dissertation_project). The code from *Team 23* [@team23] can be retrieved from their [OSF repository](https://osf.io/akqt4/). Following R packages were used *here 1.0.1* [@here], *data.table 1.14.0* [@datatable], *tidyverse 1.3.1* [@tidyverse], *lme4 1.1-27.1* [@lme4], *pbmcapply 1.5.0* [@pbmcapply], *PupillometryR 0.0.3* [@PupillometryR] and *cowplot 1.1.1* [@cowplot].

## Data description and preparation
     The data was retrieved from @silberzahn2018. It contained information about football players, their encounters with referees and the received cards (yellow, yellow/red and red). Moreover, it included a player’s position, age, club, league country, victories, ties, defeats, and goals. In addition, a skin tone rating based on two independent judges was included. The referees were numerically coded to protect their identity. The referees’ countries of origin were also included as well as implicit and explicit racism bias scores for their respective countries. The exploratory data analysis (EDA) of *Team 23* showed that each row of the dataset represents a unique player-referee combination listing all their encounters as well as the couple’s total number of received/assigned cards. *Team 23* stated that it preferred a different data format where each player-referee encounter is reflected by one row. This way, each encounter had a maximum of one red card. To achieve this format, the data had to be transformed, i.e. disaggregated. Further EDA showed that receiving a red card was highly unlikely ($p = 0.008%$), i.e. the data were highly skewed. The team, therefore, used a logistic regression which is a statistical modelling technique equipped to deal with skewed/binary distributions. Figure \ref{fig:var_plots} shows an overview of the properties of the core variables used in their analysis.

 Finally, the team excluded all referees who did not have at least 22 encounters with players. Every football game includes (at least) 22 players, 11 players per team. If a referee has less than 22 player-encounters, there are missing cases. According to team 23, this was mostly the case in referees officiating games of minor leagues. As they focused on the major European leagues (England, Germany, Italy, and Spain), referees with less than 22 player-encounters were excluded. This step excluded about 66% of the referees but retained 97.4% of all player-referee combinations. The final dataset used for the current project contained 335,537 observations and 19 variables.  \newline

```{r var_plots, echo = FALSE, fig.width = 8, fig.align = "center", fig.cap = "Team 23 base variables properties. A) There was a strong skew towards receiving no red card compared to receiving a red card. B) Most players had a lower skin tone rating, i.e., a brighter skin colour. C) The implicit racial bias scores hardly varied. D) The explicit racial bias scores varied slightly more than the implicit scores. (The latter two were centred around the mean; hence, zero does not mean there is no bias.)"}

@@

# Results
## Replicating *Team 23* analysis
     First, the analysis of *Team 23* was replicated using their scripts. Skin tone significantly affected the odds of being sent off the field ($OR = 1.311$, 95%CI [1.099, 1.563], $p = 0.003$). This means that when keeping all other variables constant, for every unit increase in skin tone rating (darker skin tone), the odds of being sent off the field increased by about 131%. The interaction terms of skin tone and implicit racial bias ($OR = 0.004$, 95%CI [0.000, 23.259], $p = 0.211$) as well as skin tone and explicit racial bias ($OR = 1.837$, 95%CI [0.493, 6.848], $p = 0.365$) were both non-significant. In accordance with the original analysis concerning additional variables, there were significant differences of OR between leagues and positions as well as implicit racial bias scores. Explicit bias scores were on the verge of being non-significant. Thus, the results of team 23 were reproducible without any adjustment.
